---
title: "Coding Assignment: Time Series and Trends"
subtitle: "A quick recap + practical extensions (line charts, smoothing, annotation, uncertainty)"
author: "Jared Edgerton"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, plsc498.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: ""
---

class: center, middle, inverse, title-slide

.title[
# Coding Assignment: Time Series and Trends
]
.subtitle[
## A quick recap + practical extensions (line charts, smoothing, annotation, uncertainty)
]
.author[
### Jared Edgerton
]

---

# What you should remember

This is a short recap and extension.

You should already be comfortable with:

- Line charts as claims about temporal order
- Trend vs seasonality vs noise
- Annotation (vertical lines, text labels)
- Smoothing as a modeling decision

Today we extend that with:

- building line charts from data with dates
- rolling means vs LOESS in practice
- adding uncertainty ribbons (`geom_ribbon`)
- formatting date axes (`scale_x_date`)
- multiple series (color vs facets)

---

# Setup

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

set.seed(42)
```

---

# Simulated data

We simulate quarterly approval data for two parties over 15 years, with: (a) different trends, (b) a seasonal component, (c) an event shock, and (d) uncertainty (standard errors).

```{r}
quarters <- seq(as.Date("2008-01-01"), as.Date("2022-10-01"), by = "quarter")
n_q <- length(quarters)

df <- expand.grid(date = quarters, party = c("Party A", "Party B")) %>%
  arrange(party, date) %>%
  group_by(party) %>%
  mutate(
    t = row_number(),
    trend = ifelse(party == "Party A", 48 + 0.15 * t, 42 - 0.05 * t),
    season = 2 * sin(2 * pi * t / 4),
    shock = ifelse(party == "Party A" & date >= as.Date("2016-01-01") &
                     date <= as.Date("2016-07-01"), -6, 0),
    noise = rnorm(n(), 0, 2),
    approval = trend + season + shock + noise,
    se = runif(n(), 1.5, 3.5),
    lower = approval - 1.96 * se,
    upper = approval + 1.96 * se,
    year = as.numeric(format(date, "%Y"))
  ) %>%
  ungroup()
```

---

# Quick check

```{r}
df %>%
  group_by(party) %>%
  summarise(
    n = n(),
    min_date = min(date),
    max_date = max(date),
    mean_approval = round(mean(approval), 1)
  )
```

---

# Starter pattern: single line chart

```{r, fig.width=7, fig.height=4}
df_a <- df %>% filter(party == "Party A")

ggplot(df_a, aes(x = date, y = approval)) +
  geom_line(linewidth = 0.6) +
  theme_classic() +
  labs(x = NULL, y = "Approval (%)",
       title = "Party A: quarterly approval")
```

---

# Starter pattern: multiple series (color)

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = date, y = approval, color = party)) +
  geom_line(linewidth = 0.6) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = NULL, y = "Approval (%)", color = NULL,
       title = "Two parties: overlaid time series")
```

---

# Starter pattern: faceted time series

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = date, y = approval)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ party, ncol = 1) +
  theme_classic() +
  labs(x = NULL, y = "Approval (%)",
       title = "Two parties: faceted for clarity")
```

---

# Smoothing: rolling mean

A rolling mean averages the last *k* observations. Transparent and easy to explain.

```{r, fig.width=7, fig.height=4}
df_a <- df_a %>%
  arrange(date) %>%
  mutate(roll_4 = zoo::rollmean(approval, k = 4, fill = NA, align = "right"))

ggplot(df_a, aes(x = date)) +
  geom_line(aes(y = approval), alpha = 0.35, linewidth = 0.4) +
  geom_line(aes(y = roll_4), color = "steelblue", linewidth = 1) +
  theme_classic() +
  labs(x = NULL, y = "Approval (%)",
       title = "Party A: raw data + 4-quarter rolling mean")
```

---

# Smoothing: LOESS

```{r, fig.width=7, fig.height=4}
ggplot(df_a, aes(x = date, y = approval)) +
  geom_line(alpha = 0.35, linewidth = 0.4) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1, color = "firebrick") +
  theme_classic() +
  labs(x = NULL, y = "Approval (%)",
       title = "Party A: raw data + LOESS smoother")
```

---

# Annotation: marking an event

```{r, fig.width=7, fig.height=4}
ggplot(df_a, aes(x = date, y = approval)) +
  geom_line(linewidth = 0.5) +
  geom_vline(xintercept = as.Date("2016-01-01"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2016-04-01"), y = max(df_a$approval) - 2,
           label = "Crisis event", hjust = 0, size = 3.5) +
  theme_classic() +
  labs(x = NULL, y = "Approval (%)",
       title = "Party A: annotated time series")
```

---

# Uncertainty ribbons

Use `geom_ribbon()` to show confidence bands around a time series.

```{r, fig.width=7, fig.height=4}
ggplot(df_a, aes(x = date)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey70", alpha = 0.4) +
  geom_line(aes(y = approval), linewidth = 0.6) +
  theme_classic() +
  labs(x = NULL, y = "Approval (%)",
       title = "Party A: approval with 95% confidence ribbon")
```

Interpretation:
- the ribbon shows the **estimated uncertainty** around each value
- wider ribbons = less certainty

---

# Uncertainty ribbons for two groups

```{r, fig.width=7, fig.height=4}
ggplot(df, aes(x = date, fill = party, color = party)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(aes(y = approval), linewidth = 0.6) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = NULL, y = "Approval (%)", color = NULL, fill = NULL,
       title = "Both parties: approval with uncertainty")
```

---

# Formatting date axes

Control date labels with `scale_x_date()`.

```{r, fig.width=7, fig.height=4}
ggplot(df_a, aes(x = date, y = approval)) +
  geom_line(linewidth = 0.5) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  theme_classic() +
  labs(x = NULL, y = "Approval (%)",
       title = "Formatted date axis (2-year breaks)")
```

---

# Year-over-year comparison

Sometimes it helps to overlay years on the same axis (month or quarter).

```{r, fig.width=7, fig.height=4}
df_recent <- df %>%
  filter(party == "Party A", year >= 2018) %>%
  mutate(quarter = as.numeric(format(date, "%m")))

ggplot(df_recent, aes(x = quarter, y = approval, color = factor(year), group = year)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = c("Q1", "Q2", "Q3", "Q4")) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = NULL, y = "Approval (%)", color = "Year",
       title = "Party A: year-over-year seasonal comparison")
```

---

# Save a figure

```{r, fig.width=7, fig.height=4}
p <- ggplot(df, aes(x = date, y = approval, color = party)) +
  geom_line(linewidth = 0.6) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = NULL, y = "Approval (%)", color = NULL)

p

ggsave("time_series_two_parties.png", plot = p, width = 7, height = 4, dpi = 300)
```
