---
title: "Coding Assignment: Uncertainty, Redesign, and Publication Quality"
subtitle: "A quick recap + practical extensions (error bars, ribbons, checklists, export)"
author: "Jared Edgerton"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, plsc498.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: ""
---

class: center, middle, inverse, title-slide

.title[
# Coding Assignment: Uncertainty, Redesign, and Publication Quality
]
.subtitle[
## A quick recap + practical extensions (error bars, ribbons, checklists, export)
]
.author[
### Jared Edgerton
]

---

# What you should remember

This is a short recap and extension.

You should already be comfortable with:

- Error bars and confidence intervals
- Uncertainty ribbons on line charts
- The publishing-quality checklist
- Color accessibility basics

Today we extend that with:

- building error bar plots from summary data
- adding ribbons to time series
- redesigning a "before" plot into a polished "after" version
- export settings (dpi, size, format)
- peer critique workflow

---

# Setup

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

set.seed(42)
```

---

# Simulated survey data

We simulate survey estimates for policy support across countries and policy areas.

```{r}
df_survey <- expand.grid(
  country = c("Aurelia", "Belvica", "Cordonia", "Deltara"),
  policy = c("Climate", "Immigration", "Healthcare")
) %>%
  mutate(
    estimate = round(runif(12, 30, 75), 1),
    n = sample(400:1200, 12, replace = TRUE),
    se = round(sqrt(estimate * (100 - estimate) / n), 2),
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se
  )
```

---

# Quick check

```{r}
df_survey %>%
  select(country, policy, estimate, se, n) %>%
  arrange(policy, desc(estimate))
```

---

# Starter pattern: dot plot with error bars

```{r fig.height=4, fig.width=7, warning=FALSE, message=FALSE}
df_climate <- df_survey %>% filter(policy == "Climate") %>%  mutate(country = reorder(country, estimate))

ggplot(df_climate, aes(x = estimate, y = country)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.25) +
  geom_point(size = 3) +
  theme_classic() +
  labs(x = "Support (%)", y = NULL, title = "Climate policy support with 95% CIs")
```

---

# Starter pattern: grouped error bars

```{r, fig.width=7, fig.height=5}
df_survey <- df_survey %>% mutate(country = factor(country))

ggplot(df_survey, aes(x = estimate, y = country, color = policy)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.25, position = position_dodge(0.6)) +
  geom_point(size = 2.5, position = position_dodge(0.6)) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = "Support (%)", y = NULL, color = NULL, title = "Policy support by country and topic")
```

---

# Simulated time series with uncertainty

```{r}
df_ts <- data.frame(
  year = rep(2005:2024, 2),
  party = rep(c("Party A", "Party B"), each = 20)
) %>%
  group_by(party) %>%
  mutate(
    t = row_number(),
    estimate = ifelse(party == "Party A",
                      48 + 0.4 * t + rnorm(20, 0, 2),
                      52 - 0.2 * t + rnorm(20, 0, 2.5)),
    se = runif(20, 1.5, 4),
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se
  ) %>%
  ungroup()
```

---

# Time series with uncertainty ribbons

```{r, fig.width=7, fig.height=4}
ggplot(df_ts, aes(x = year, color = party, fill = party)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, color = NA) +
  geom_line(aes(y = estimate), linewidth = 0.7) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = NULL, y = "Approval (%)", color = NULL, fill = NULL,
       title = "Party approval with 95% confidence ribbons")
```

---

# The redesign exercise

Now we practice the **publication pass**. Take a rough figure and make it polished.

Here is a "before" version (deliberately rough):

```{r, fig.width=7, fig.height=4}
p_before <- ggplot(df_climate, aes(x = estimate, y = country)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower, xmax = upper))

p_before
```

---

# Redesign: applying the checklist

Apply each checklist item one at a time:

1. Clear title
2. Axis labels with units
3. Caption with source
4. Reference line (e.g., 50% = majority)
5. Color for emphasis
6. Readable text size
7. Clean theme

---

# After the checklist

```{r, fig.width=7, fig.height=4, message=FALSE, warning=FALSE}
p_after <- ggplot(df_climate, aes(x = estimate, y = country)) +
  geom_vline(xintercept = 50, linetype = "dashed", alpha = 0.4) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.2, color = "grey50") +
  geom_point(size = 3, color = "steelblue") +
  theme_classic(base_size = 13) +
  labs(x = "Support (%)", y = NULL, title = "Public support for climate policy", subtitle = "Point estimates with 95% confidence intervals", caption = "Source: Simulated survey data | Dashed line = 50% threshold")
p_after
```

---

# Export: saving at publication quality

```{r}
ggsave("climate_support_polished.png",
       plot = p_after,
       width = 7, height = 4,
       dpi = 300, units = "in")
```

Key settings:
- `dpi = 300` for print-quality resolution
- `width` and `height` control aspect ratio explicitly
- PNG for raster, PDF for vector graphics

---

# Peer critique workflow

When reviewing a classmate's figure:

1. **What is the main message?** (Can you state it in one sentence?)
2. **What is mapped vs styled?** (Is the encoding appropriate?)
3. **Is uncertainty shown?** (If applicable)
4. **Checklist items:** title, labels, units, caption, legend, color, aspect ratio
5. **One concrete suggestion** (not "make it better" but "move the legend to bottom")

---

# Color accessibility check

```{r, fig.width=7, fig.height=4}
ggplot(df_survey, aes(x = estimate, y = country, color = policy, shape = policy)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.25, position = position_dodge(0.6)) +
  geom_point(size = 2.5, position = position_dodge(0.6)) +
  scale_color_brewer(palette = "Dark2") +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = "Support (%)", y = NULL, color = NULL, shape = NULL,
       title = "Redundant encoding: color + shape for accessibility")
```

---

# Shiny starter (for your final project)

A minimal Shiny app for one of today's figures:

```r
library(shiny)
library(ggplot2)
library(dplyr)

ui <- fluidPage(
titlePanel("Policy Support Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("policy", "Choose a policy area:",
                  choices = c("Climate", "Immigration", "Healthcare"))
    ),
    mainPanel(plotOutput("supportPlot"))
  )
)

```

---

# Shiny starter (for your final project)


```r

server <- function(input, output) {
  output$supportPlot <- renderPlot({
    df_filtered <- df_survey %>% filter(policy == input$policy) %>%
      mutate(country = reorder(country, estimate))
    ggplot(df_filtered, aes(x = estimate, y = country)) +
      geom_vline(xintercept = 50, linetype = "dashed", alpha = 0.4) +
      geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
      geom_point(size = 3, color = "steelblue") +
      theme_classic(base_size = 14) +
      labs(x = "Support (%)", y = NULL,
           title = paste(input$policy, "policy support"))
  })
}

shinyApp(ui = ui, server = server)
```

---

# Save a figure

```{r, fig.width=7, fig.height=4}
p <- ggplot(df_ts, aes(x = year, color = party, fill = party)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, color = NA) +
  geom_line(aes(y = estimate), linewidth = 0.7) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = NULL, y = "Approval (%)", color = NULL, fill = NULL)

p

ggsave("approval_uncertainty.png", plot = p, width = 7, height = 4, dpi = 300)
```
