---
title: "Coding Assignment: Geospatial Data and Mapping"
subtitle: "A quick recap + practical extensions (choropleths, sf, joins, projections)"
author: "Jared Edgerton"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, plsc498.css]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: ""
---

class: center, middle, inverse, title-slide

.title[
# Coding Assignment: Geospatial Data and Mapping
]
.subtitle[
## A quick recap + practical extensions (choropleths, sf, joins, projections)
]
.author[
### Jared Edgerton
]

---

# What you should remember

This is a short recap and extension.

You should already be comfortable with:

- Choropleths as fill-colored geographic regions
- Counts vs rates (normalization)
- Sequential, diverging, and categorical color scales
- The ecological fallacy

Today we extend that with:

- building maps using `map_data()` and `sf` objects
- joining attribute data to boundary data
- projection choices with `coord_sf()`
- binning and legend design for clarity
- when to use a map vs a non-spatial chart

---

# Setup

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(sf)
library(maps)
library(scales)

set.seed(42)
```

---

# Map data: two approaches

**Approach 1: `map_data()` from the `maps` package**

Returns a data frame of polygon coordinates. Simple but limited.

```{r}
us <- map_data("state")
head(us, 4)
```

**Approach 2: `sf` objects**

The modern standard. Geometries are stored in a special column. Supports projections, spatial operations, and easy ggplot integration.

---

# Starter pattern: basic US map with map_data

```{r, fig.width=7, fig.height=4}
ggplot(us, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "grey90", color = "white", linewidth = 0.3) +
  coord_quickmap() +
  theme_void() +
  labs(title = "US states (map_data approach)")
```

---

# Starter pattern: sf approach (world map)

```{r, fig.width=7, fig.height=4}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

ggplot(world) +
  geom_sf(fill = "grey90", color = "white", linewidth = 0.2) +
  theme_void() +
  labs(title = "World map (sf approach with rnaturalearth)")
```

---

# Building a choropleth: step by step

1. Get boundary data (polygons with a region identifier)
2. Get your attribute data (one row per region)
3. Make sure the join key matches (e.g., both lowercase state names)
4. Join
5. Map the variable to `fill`

---

# Simulated political data

```{r}
state_data <- data.frame(
  region = tolower(state.name),
  dem_share = rnorm(50, mean = 50, sd = 8),
  median_income = round(rnorm(50, mean = 55000, sd = 10000)),
  pop = state.x77[, "Population"] * 1000
) %>%
  mutate(
    dem_share = pmin(pmax(dem_share, 25), 75),
    dem_win = ifelse(dem_share > 50, "Dem", "Rep")
  )
```

---

# Join and map: continuous variable

```{r, fig.width=7, fig.height=4}
us_merged <- us %>%
  left_join(state_data, by = "region")

ggplot(us_merged, aes(x = long, y = lat, group = group, fill = dem_share)) +
  geom_polygon(color = "white", linewidth = 0.3) +
  coord_quickmap() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue",
                       midpoint = 50, name = "Dem %") +
  theme_void() +
  labs(title = "Simulated two-party vote share (diverging scale)")
```

---

# Join and map: categorical variable

```{r, fig.width=7, fig.height=4}
ggplot(us_merged, aes(x = long, y = lat, group = group, fill = dem_win)) +
  geom_polygon(color = "white", linewidth = 0.3) +
  coord_quickmap() +
  scale_fill_manual(values = c("Dem" = "steelblue", "Rep" = "firebrick"),
                    name = "Winner") +
  theme_void() +
  labs(title = "Simulated election result (categorical)")
```

---

# Projections matter

```{r, fig.width=7, fig.height=4}
us_sf <- st_as_sf(map("state", plot = FALSE, fill = TRUE))

ggplot(us_sf) +
  geom_sf(fill = "grey90", color = "white", linewidth = 0.3) +
  coord_sf(crs = st_crs(5070)) +
  theme_void() +
  labs(title = "US states: Albers equal-area projection (EPSG:5070)")
```

Interpretation:
- Albers preserves area: states are proportional to their real size
- Better for choropleths than the default Mercator-like projection

---

# Binning for choropleths

Binning a continuous variable into classes can make the map easier to read.

```{r, fig.width=7, fig.height=4}
state_data <- state_data %>%
  mutate(income_cat = cut(median_income,
    breaks = c(0, 45000, 55000, 65000, Inf),
    labels = c("< $45k", "$45k-$55k", "$55k-$65k", "> $65k")))

us_merged <- us %>%
  left_join(state_data, by = "region")

ggplot(us_merged, aes(x = long, y = lat, group = group, fill = income_cat)) +
  geom_polygon(color = "white", linewidth = 0.3) +
  coord_quickmap() +
  scale_fill_brewer(palette = "YlGnBu", name = "Median income") +
  theme_void() +
  labs(title = "Median income by state (binned)")
```

---

# World choropleth: life expectancy

```{r, fig.width=7, fig.height=4}
gap_2007 <- gapminder::gapminder %>%
  filter(year == 2007)

world_merged <- world %>%
  left_join(gap_2007, by = c("name" = "country"))

ggplot(world_merged) +
  geom_sf(aes(fill = lifeExp), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "C", name = "Life exp.", na.value = "grey80") +
  theme_void() +
  labs(title = "Life expectancy by country (2007)")
```

---

# When a map is not the best chart

Sometimes a dot plot or bar chart communicates the same information more precisely.

```{r, fig.width=6, fig.height=4}
top_states <- state_data %>%
  arrange(desc(median_income)) %>%
  slice_head(n = 15) %>%
  mutate(region = reorder(region, median_income))

ggplot(top_states, aes(x = median_income, y = region)) +
  geom_point(size = 2.5) +
  scale_x_continuous(labels = dollar) +
  theme_classic() +
  labs(x = "Median income", y = NULL,
       title = "Top 15 states by median income (dot plot)")
```

---

# Save a figure

```{r, fig.width=7, fig.height=4}
p <- ggplot(us_merged, aes(x = long, y = lat, group = group, fill = dem_share)) +
  geom_polygon(color = "white", linewidth = 0.3) +
  coord_quickmap() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue",
                       midpoint = 50, name = "Dem %") +
  theme_void()

p

ggsave("choropleth_vote_share.png", plot = p, width = 7, height = 4, dpi = 300)
```
